---
title: "Exercise 02"
author: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    always_allow_html: yes
    toc: yes
    number_sections: yes
    code_folding: hide
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this homework, you will save the `R Markdown` (e.g., `.Rmd`) homework document and save it to your `homework` project directory on your computer. Save a new copy of this file so that its name has the prefix `2023_Lastname_FirstInitial`. Full name example: `2023_cookg_EX_02.Rmd`. 
Feel free to work with a partner but understand that the work submitted has to be your own.

**This homework is due on Tuesday by 2:30pm.**


*Enter your name here*   



**Problem 1: Checking your working directory using `{here}`**

You should have the **{here}** library installed. The following code block contains a function from the **{here}** library that will return the name of the your working directory. If the function does not run, you will need to install **{here}**. In addition, if the working directory is not put in a place where your homework goes, you will need to either move your `.Rmd` homework file or set up a project directory for homework.

```{r}
here::here()
```



**Problem 2: Loading libraries** 

Inside the following code block, use `library()` to load the **{readr}** and **{dplyr}** libraries. You won't use them for this exercise but you will for the next one.

```{r, libraries}

```

**Problem 3: Creating objects** 

Create a code chunk in R Markdown by either a) typing on a Windows system `CONTROL+ALT+i` or on a Mac `COMMAND+OPTION+i` all at the same time or by 2) clicking the green icon containing a + and a c that appears in RStudio near the file tab for your opened `.Rmd` file. You can also [go here](https://rmarkdown.rstudio.com/lesson-3.html) to see the icon image. 

In that code chunk: 

1) assign the character string "hello" to an object named x, 
2) assign any numeric object to y

```{r}


```


**Problem 4: Adding code snippets to RStudio**

Code snippets are useful for creating shortcuts for task you repeat. Copy the snippet below and then got to `Tools -> Edit Code Snippets` and paste the snippet. Save. 

```
snippet proj_dir
	suppressWarnings(lapply(c("data", "r", "figs", "report", "refs", "notes", "notes/more_notes"), FUN = function(x) dir.create(here::here(x))))
```



**Problem 5: Running code snippets**

For this snippet, you only have to do it once, so the recommendation is to type the snippet hotstring `proj_dir` at the `R` console prompt. When you start typing, RStudio will likely start to auto-populate options and one that will appear is `proj_dir   {snippet}`. Select it an run that code in the console.

The snippet is using the `lapply()` function to apply a function to either a `list` or a `vector` object. The `vector` is a character vector containing several elements which will become names of directories (e.g., `"/r"`) or sub-directories (e.g., `"/notes/more notes"`. That `vector` is then passed to `dir.create()`, which will create (iteratively, element by element) a directory (or sub-directory) at the specified path location. That location is the path returned by `here::here()`. 


Run the code snippet at the console to create the directories. Must be done for next step to work.




**Problem 6: Creating a `.R` Script file**

`.R` files are different from `.Rmd` files. They contain only `R` code. In `RStudio`, create a new file that is an `R Script` file. `Save As` `my_functions` to your newly created `.../homework/r` directory (previous problem). The file extension will be `.R` so your file will now be `my_functions.R`. In that file, paste the code below, which is a function I wrote for viewing data frames and tibbles in `html` form. I prefer this function over the built-in `View()` function. Maybe you won't prefer it but this is homework anyway.


```
view_html <- function(object, rows = F, show = 100,...) { 
    
    if (!require(DT)) {
        
        stop("DT library not installed")
        
    } else {
        
        if (tibble::is_tibble(object)) {
            object = as.data.frame(object)
            #message("converted tibble to dataframe for viewing")
        } 
        
        if (is.null(dim(object)) & class(object) == "list") {
            message("Object is a list. Viewer displays last list element. Consider passing each element to view().")
            
            lapply(object, function(x) {
                DT::datatable(x, rownames = rows, options = list(pageLength = show))
            }) 
        } else {
            DT::datatable(object, rownames = rows, options = list(pageLength = show)) 
        }
        
    }
}

```

Close the `my_functions.R` file.



**Problem 7: Sourcing `.R` Script files**

Often times, you will have code that you do not want to add to your `RMarkdown` file because that file is likely more for reports than for running raw code. You don't want that file getting too busy. Within the `RMarkdown` file, you can use `source()` to call and execute code saved in a different file (e.g., `my_functions.R`)

Sourcing will require you to use `here::here()` because finding the path location will be easy (if it exists). The function in following code block will return a character object of the `...path to project working dir/r/my_functions.R"` 

```{r}
here::here("r", "my_functions.R")
```

You can test whether the full path exists using `file.exists()`.

```{r}
file.exists(here::here("r", "my_functions.R"))
```

In this final code block, `source()` that file path (hint: you want the path only, you do not want to know whether the path exists). The first argument in `source()` is `file`, so the `file` is that path. Type ?source in the R console to see the help file if you want to read more. 

```{r}

```

Having sourced the file, the `view_html()` function is now defined and running in R. 



**Problem 8: Viewing data frames/tibbles**

To test out the difference between `View()` and `view_html()`, you will use them on the `mtcars` data frame built into `R`. Make note of the approaches to viewing the data frame. To shorted things up, we will wrap the function with `head()` so that only the first few rows of data appear. 

```{r}
head(mtcars)

View(head(mtcars))
```


In this code block, pass `mtcars` to `view_html()` to view with that function. 

```{r, screenshot.force = TRUE}


```



**Problem 9: RMarkdown headings**

Practice using `RMarkdown` to create headers for sections. Create 3 headers (e.g., level 1, level 2, and level 3). You can find tips [here](https://rmarkdown.rstudio.com/authoring_basics.html). Once you render you `html` file, take note of what these headers do.




**Problem 10: RMarkdown lists**

Practice using some `RMarkdown` code by creating: **a)** bulleted list with three items that describe you, **b)** a numbered list with your top 3 band names, and **c)** a sentence that has one word in bold and one word in italics. You can find some tips [here](https://rmarkdown.rstudio.com/lesson-8.html).




**Rendering RMarkdown files for submission**

You will want to produce two rendered documents. You can do this using code or more simply using the toolbar icon. You can find tips for both approaches [here](https://rmarkdown.rstudio.com/lesson-9.html).

However, I have discovered that if there are `html` objects present in your `RMarkdown` file, they will not easily render to `pdf` format. A solution is to install a library named `webshot` (so install it) and then install PhantomJS using `webshot::install_phantomjs()` and the console. You may still see an error but the file should be created. If you run into error, just submit `html`.  

In order to render `pdf` files from `RMarkdown`, you will need a `LaTeX` installation. If you use `LaTeX`, you likely already have this set up. If not, when you `knit` from the toolbar and select `knit to pdf`, `R` will throw an error and suggest a simple installation approach which is to execute `tinytex::install_tinytex()`. So, run that code in your `R` console and then `knit` your homework as a `pdf` so that you can see what it looks like and so that you can compare that output to the `html` version.



**Please knit an `html` file and a `pdf` file and upload both [here](https://claremontmckenna.app.box.com/f/140969a32dea44e9ab6c0c00146d66e3).** 
