---
title: "Exercise 01"
author: "yourname"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  #word_document:
  #  toc: yes
  #  number_sections: yes
  html_document:
    toc: no
    number_sections: yes
    code_folding: show #hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this homework, you will save the `R Markdown` (e.g., `.Rmd`) homework document and save it to your `homework` project directory on your computer. Save a new copy of this file so that its name has the prefix `2023_Lastname_FirstInitial`. Full name example: `2023_cookg_EX_09.Rmd`. Feel free to work with a partner but understand that the work submitted has to be your own.

This homework utilizes examples from these modules:

- ****


**This homework is due on Tuesday by 2:30pm.**


*Enter your name in the author YAML code above*   


```{r}
source(here::here("r", "my_functions.R"))
```

# (1pt) **Load only used libraries**

```{r echo=TRUE}

```

**Load data**

Some questions will utility the data below. 

```{r}
DAT <- readr::read_csv("https://github.com/slicesofdata/fods24/raw/main/data/?.csv", show_col_types = F)
```

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****

# (1pt) ****


**Please knit an `html` file and upload [here](https://claremontmckenna.app.box.com/f/140969a32dea44e9ab6c0c00146d66e3).** 

